---
title: "Markov Chain Example"
author: "Marco Smolla"
date: "3/22/2021"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is a Markov Chain? 
Markov Chains describe the transition probabilities of a system from one state to another. For example, if a system has the state 'activated', 'stand-by', and 'deactivated', what is the probability to go from 'stand-by' to 'activated', and so on. For more information, make sure to take a look at the corresponding [Wikipedia entry](https://en.wikipedia.org/wiki/Markov_chain) and these [lecture notes](https://www.stat.auckland.ac.nz/~fewster/325/notes/ch8.pdf). 

## Which packages do we need?
There are a bunch of good `R` packages out there. Here, I am relying on `markovchain` and for plotting I will use `igraph`. 

```{r}
# For calculating markov chains 
library(markovchain) 
# For plotting as a network
library(igraph) 
```

## Example data
Here is our example data:
```{r}
# Example transitions for a single individual (H=helper, R=reciprocator, B=beneficiary)
x <- sample(c("H", "R", "B"), 1000, replace = T, prob=c(.30, .60, .10)) # here I am creating a random vector of states (which in your case would be the sequence of states of your individual buyers)

# Count: 
table(x)

# Proportion
table(x)/length(x)
```

## Calculating a transition matrix - the packaged way
If you prefer to use a ready-made function, use the `markovchainFit()` function from the `markovchain` package: 
```{r}
# using markovchainFit from the markovchain package
mcFit <- markovchainFit(data=x)
# look at results
show(mcFit)
```

# Calculating a transition matrix - DIY
Alternatively, we can build our own `trans.matrix()` function:
```{r}
trans.matrix <- function(X, prob=T)
{
 # turn sequence into a factor 
 if(!is.factor(x)) x <- as.factor(x)
 # define individual levels of the factor
 ids <- levels(x)
 # create an empty ids x ids matrix
 m <- matrix(0, ncol=length(ids), nrow=length(ids))
 # turn sequence into levle numbers (1, 2, 3, ...)
 state <- as.numeric(x)
 # create a two column matrix where the first column are t and the second column t+1 entries 
 tmp <- cbind(state[-length(state)], state[-1])
 # add up all individual transitions (i.e. at t the state is 1 and at t+1 the state is 2, then add 1 to the number in the matrix M_12)
 for(i in 1:nrow(tmp)){
  m[tmp[i,1], tmp[i,2]] <- m[tmp[i,1], tmp[i,2]] + 1
 }
 # turn counts into probabilities
 if(prob) m <- m / rowSums(m)
 # turn matrix into data.frame with column names and row names 
 df <- data.frame(m, row.names = ids)
 colnames(df) <- ids
 # return transitio matrix
 return(df)
}
```

Let us test this function without initial sequence `x`: 

```{r}
# calculate transition matrix
mcFit2 <- trans.matrix(x)
mcFit2
```

Compare `mcFit2` with the estimate of `mcFit` and you'll see that this produces the same result. Of course, when using the markovchain package you receive much more information than the transition matrix.

## Plot transition matrix as a graph
Finally, let us plot the result as an `igraph` network:
```{r}
# turn transition matrix into a network
net <- graph_from_adjacency_matrix(adjmatrix = as.matrix(mcFit2), weighted = T)
# plot transiton matrix as network
plot(net, 
     edge.width=E(net)$weight*5, # use transition probabilities to vary edge widths
     edge.curved=T, # make edges curved (instead of straight)
     # vertex.size=50, # this is for fixed vertex size
     vertex.size=colSums(mcFit2)*50, # use in-edge weights (i.e. column sums of the transition matrix) to set the size of a vertex (larger vertices have larger weights of incoming edges)
     edge.label=round(E(net)$weight, 2)) # add transition probabilities as labels to all edges (and round values to make it easier to read)
```

